Compositional Quantitative Resource Analysis of C Programs

### 2014-06-20 paper/main.tex
add example derivation
more work on the intro

In software engineering and software verification, we often would like
to have static information about the quantitative behavior of
programs.  For example, stack and heap-space bounds are important to
ensure the reliability of safety-critical systems [XXX (see PLDI)].
% Clock-cycle bounds are needed to guarantee the safety of real-time
% systems [XXX wilhelm].  
Static energy usage information is critical for autonomous systems and
has applications in cloud computing [XXX?  (cite this guy form the
us))].  Worst-case resource bounds can be used to prevent side-channel
attacks through resource padding [XXX?].  Loop and
recursion-depth bounds are used to ensure the accuracy of programs
that are executed on unreliable hardware [XXX oopsla] and complexity
bounds are needed to verify security protocols [XXX].
% XXX Mention differential privacy here?
%
In general, quantitative performance information at design time can
provide useful feedback for developers.

@@

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
notations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
To understand how the analysis works, it is instructive to look at
some more involved examples.  Examples \emph{speed\_1} and
\emph{speed\_2} are taken from Gulwani et al~\cite{Gulwani-speed}.
They are both handled by the SPEED tool~\cite{Gulwani-speed} but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

@@

\begin{figure}
\begin{lstlisting}
${\color{blue} \{ A(0, N) + N \} }$
i=1; m=a[0];
${\color{blue}%
  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-i) \} }$
while (i<N) {
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
  if (a[i] > m)
    m = a[i];
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1, N) + (N-i) \} }$
  i = i+1;
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
}
${\color{blue} \{ 0 \} }$
\end{lstlisting}
\caption{Example derivation where we noted $A(i, N)$
  for $\#\{ k \mid k \ge i \land \forall\, 0\le j<k.\, a[j] < a[k]\}$}
\label{fig:exemple}
\end{figure}

@@

Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
al~\cite{Gulwani-speed}.  They are both handled by the SPEED tool but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{fig2\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{fig2\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

>>>>

In software engineering and software verification, we often would like
to have static information about the quantitative behavior of
programs.  For example, stack and heap-space bounds are important to
ensure the reliability of safety-critical systems~\cite{veristack14,Regehr05}.
% Clock-cycle bounds are needed to guarantee the safety of real-time
% systems [XXX wilhelm].  
Static energy usage information is critical for autonomous systems and
has applications in cloud computing~\cite{CohenZSL12,CarrollH10}.
Worst-case time bounds can help to create constant-time
implementations that prevent side-channel
attacks~\cite{KasperS09,BartheBCLP14}.  Loop and recursion-depth
bounds are used to ensure the accuracy of programs that are executed
on unreliable hardware~\cite{CarbinMR13} and complexity bounds are
needed to verify cryptography protocols~\cite{BartheGB09}.
% XXX Mention differential privacy here?
%
In general, quantitative performance information at design time can
provide useful feedback for developers.

@@

Note that our notion of a potential function is a refinement of the
concept of a ranking function.  A potential function can be used like
a ranking function if we use the tick metric and add the statement
\code{tick(1)} to every back edge of the program (loops and function
calls).  However, a potential function is more flexible.  For example,
we can use a potential function to prove that Example~2 does not
consume any resources in the tick metric.
\begin{lstlisting}
  while (x<y) {tick(-1); x=x+1; tick(1);} #\hfill \textnormal{(Example 2)}#
\end{lstlisting}
Similarly we can prove that Example~3 can be bounded by $10|[x,y]|$.
In both cases, we reason exactly like in the first version of the
while loop to prove the bound.  Of course, such loops with different
notations can be seamlessly combined in a larger program.
\begin{lstlisting}
  while (x<y) {x=x+1; tick(10);} #\hfill \textnormal{(Example 3)}#
\end{lstlisting}
%
To understand how the analysis works, it is instructive to look at
some more involved examples.  Examples \emph{speed\_1} and
\emph{speed\_2} are taken from Gulwani et al~\cite{GulwaniMC09}.
They are both handled by the SPEED tool~\cite{GulwaniMC09} but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{speed\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{speed\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.

@@

\begin{figure}
\begin{lstlisting}
${\color{blue} \{ A(0, N) + N \} }$
i=1; m=a[0];
${\color{blue}%
  \{ (i {=} 1 \land m=a[0]) + A(1, N) + (N-1) \} }$
while (i<N) {
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
  if (a[i] > m)
    m = a[i];
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i]} a[k]) + A(i+1, N) + (N-i) \} }$
  i = i+1;
  ${\color{blue}%
    \{ (m = \max_{k \in [0, i-1]} a[k]) + A(i, N) + (N-i) \} }$
}
${\color{blue} \{ 0 \} }$
\end{lstlisting}
\caption{Example derivation where we noted $A(i, N)$
  for $\#\{ k \mid i \le k \le N \land \forall\, 0\le j<k.\, a[j] < a[k]\}$}
\label{fig:exemple}
\end{figure}

@@

Examples \emph{fig2\_1} and \emph{fig2\_2} are taken from Gulwani et
al~\cite{GulwaniMC09}.  They are both handled by the SPEED tool but
require inference of a \emph{disjunctive invariant}.  In the abstract
interpretation community, these invariants are known to be notoriously
difficult to handle.
%
In example \emph{fig2\_1} we have one loop that first increments
variable $y$ up to $m$ and then increments variable $x$ up to $n$.  We
derive the tight bound $|[x, n]| + |[y, m]|$.
%
Example \emph{fig2\_2} is more tricky and trying to understand how it
works may be challenging.  However, with the amortized analysis in
mind, using the potential transfer reasoning, it is almost trivial to
prove a bound.  While the SPEED tool has to find a fairly
involved invariant for the loop, our tool is simply reasoning locally
and works without any clever tricks. We obtain the tight bound $|[x,
n]| + |[z, n]|$.
